---
title: "Coreid_antennae_evolution"
author: "Polli & Palaoro"
date: "Last compiled on `r format(Sys.time(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 6)

library(ape)
library(phytools)
library(geiger)
library(diversitree)
library(tidyverse)
library(pander)

```

## LOADING PHYLOGENY AND DATA FILE ####

```{r}

tt = read.tree(file = './phylo/50p.treefile')

tips = read.csv(file = './data/25-10-25_antena_coreidae_summarised_2.csv', h =T, sep = ';')

#### HANDLING THE NAMES ON THE TIPS OF THE TREE AND ON OUR DATA FILE ####

rownames(tips) = tips$Species_Name

tt$tip.label = tips$Species_Name

name.check(tt, tips)
```

## LOADING ADULTS (SPECIES), ADULTS (GENUS) AND NYMPHS (GENUS) DATA ####

```{R}
sp_adults = read.csv("./data/18-08-26_Coreidae_Adults.csv", sep = ";") %>% 
  mutate(Species_Name = case_when(
    Species_Name == "Thasus_neocalifornicus" ~ "Pachylis_neocalifornicus",
    Species_Name == "Thasus_sp." ~ 'Pachylis_sp.',
    TRUE ~ Species_Name
  ))

gen_adults = read.csv("./data/18-08-26_Adults_genera.csv", sep = ",") %>% 
  mutate(Species_Name = case_when(
    Species_Name == "Thasus_neocalifornicus" ~ "Pachylis_neocalifornicus",
    Species_Name == "Thasus_sp." ~ 'Pachylis_sp.',
    TRUE ~ Species_Name
  ))

gen_nymphs = read.csv('./data/18-08-26_Nymphs_genera.csv', sep = ',') %>% 
  mutate(Species_Name = case_when(
    Species_Name == "Thasus_neocalifornicus" ~ "Pachylis_neocalifornicus",
    Species_Name == "Thasus_sp." ~ 'Pachylis_sp.',
    TRUE ~ Species_Name
  ))
```


## ROOTING THE TREE

To get a result closer to what was presented in the paper, and to be able to date the tree, it first needs a root following the information on Forthman et al. (2024). In the paper, Jadera is the root. So, let's root it.
```{r, fig.width = 10, fig.height = 14}

t1 = root(tt, node = 308, resolve.root = T)

plot.phylo(ladderize(t1))

```

### ADULTS ANALYSES AT SPECIES LEVEL ####

```{r, fig.width = 10, fig.height = 14}

# REMOVING ZEROES FROM DATASET AND DROPPING UNNECESSARY TIPS ####

# Since our goal is analyzing trait evolution in Coreidae,
# we removed the non-Coreidae species in the tree.

# We also removed tips with no data or at species level,
# labeled as "0"

adults_data = sp_adults %>% 
  filter(Antennae != 0, 
         !Species_Name %in% c("Jadera_haematoloma", "Harmostes_serratus",
                              "Oedancala_sp.", "Oncopeltus_fasciatus",
                              "Largus_sp.", "Dysdercus_mimus",
                              "Dysdercus_suturellus", "Halyomorpha_halys"))
# Now we prune the tree to the data we have on Coreidae adults.

t_adults = keep.tip(t1, adults_data$Species_Name)

```

## THE MK MODEL: ASSESSING CHANGES IN ANTENNAL TRAIT STATE 

Before running the analyses, we need to create a matrix so phytools can read our data.

```{r}
adults.sp.sim = setNames(as.factor(adults_data$Sum), adults_data$Species_Name)

adults.sp.sim
```

With the matrix done, we can now move to the analyses.

We are using the fitMk function. The function requires us to specify the tree we are going to use, the data in matrix format, the evolutionary model, and, if we wish, the state of the root. 

We are fitting two types of evolutionary models: Equal rates and All rates differ. We are not using the symmetrical because we only have two trait states. When that happens, the symmetrical evolutionary model is equal to the all rates differ model.

Lastly, we are using two types of root estimation. One where both states have equal probabilities, and another where probabilities are calculated based on the data (fitzjohn). We used both because we have no a priori hypothesis of the ancestral state.

```{r, fig.width = 5, fig.height = 5}
### NOTE: ONCE AGAIN I AM COMMENTING THE FUNCTION SO YOU DO NOT MISTAKENLY
### RUN IT. IF YOU WANT TO RUN IT, JUST UNCOMMENT THE FUNCTIONS. 
### OTHERWISE, LOAD THEM AND CARRY ON WITH THE ANALYSES.

#fit.ER.equal = fitMk(ladderize(t_adults), adults.sp.sim , model = "ER",
#                     pi = 'equal')


#fit.ER.fitz = fitMk(ladderize(t_adults), adults.sp.sim , model = "ER",
#                    pi = 'fitzjohn')

#plot(fit.ER.fitz)

#save(fit.ER.equal, file = './evo.models/er-equal-adults.RData')
#save(fit.ER.fitz, file = './evo.models/er-fitz-adults.RData')

load( file = './evo.models/er-equal-adults.RData')
load( file = './evo.models/er-fitz-adults.RData')

# All rates differ model #

#fit.ARD.equal = fitMk(ladderize(t_adults), adults.sp.sim., model = "ARD",
#                      pi = 'equal')

#fit.ARD.fitz = fitMk(ladderize(t_adults), adults.sp.sim, model = "ARD",
#                     pi = 'fitzjohn')

#save(fit.ARD.equal, file = './evo.models/ard-equal-adults.RData')
#save(fit.ARD.fitz, file = './evo.models/ard-fitz-adults.RData')

load(file = './evo.models/ard-equal-adults.RData')
load(file = './evo.models/ard-fitz-adults.RData')

# We are also using the HRM models for thoroughness. Maybe there is an
# underlying mechanism for changing antennal state we are not capturing. 

#fitHRM.er.equal = fitHRM(ladderize(t_adults), adults.sp.sim , model = "ER",
#                         pi = 'equal')

#fitHRM.er.fitz = fitHRM(ladderize(t_adults), adults.sp.sim , model = "ER",
#                        pi = 'fitzjohn')

#save(fitHRM.er.equal, file = './evo.models/hrm-er-equal-adults.RData')
#save(fitHRM.er.fitz, file = './evo.models/hrm-er-fitz-adultso.RData')

load(file = './evo.models/hrm-er-equal-adults.RData')
load(file = './evo.models/hrm-er-fitz-adults.RData')

# All rates differ HRM #

#fitHRM.equal = fitHRM(ladderize(t_adults), adults.sp.sim , model = "ARD",
#                      pi = 'equal')


#fitHRM.fitz = fitHRM(ladderize(t_adults), adults.sp.sim , model = "ARD",
#                     pi = 'fitzjohn')

#save(fitHRM.equal, file = './evo.models/hrm-equal-adults.RData')
#save(fitHRM.fitz, file = './evo.models/hrm-fitz-adults.RData')


load(file = './evo.models/hrm-equal-adults.RData')
load(file = './evo.models/hrm-fitz-adults.RData')

# Now we compare the models using AIC. 
# We chose the model with the lowest AIC as the best model.
# If models had similar AICs, we chose the one with the lowest
# degrees of freedom. In the case, below, the ER model was the best.
# Between the two root states, we used the fitjohn approach because it 
# had more AIC weight when compared to the equal ER model.

fit_models = anova(fit.ER.equal.ant, fit.ARD.equal.ant,
                    fitHRM.er.equal.ant, fitHRM.equal.ant,
                    fit.ER.fitz.ant, fit.ARD.fitz.ant,
                    fitHRM.er.fitz.ant, fitHRM.fitz.ant)

# It was chosen the ER fitz model, once it had the lowest AIC with low 
# degree of freedom = 1

anc.ant = ancr(fit.ER.fitz.ant)

fit.ER.fitz.ant
plot(fit.ER.fitz.ant)

cols = setNames(c("#0E3768","#94129F", "#de627a", "#ce4326"),
               levels(adults.sp.sim))


#pdf(file = './figures/ancestral-tree-adults-species.pdf', h = 24, w = 20)
plot(anc.ant,  
     args.plotTree = list(lwd = 3, fsize = 1, 
                          ftype = 'i',
                          outline = T,
                          cex = 0.2, offset = 1
     ),
     args.nodelabels = list(cex = 0.3, piecol = cols),
     args.tiplabels = list(cex = 0.3)
     
)
#dev.off()


```


```{r, fig.width = 10, fig.height = 15}
# Re-running the stochastic mapping takes a long while and its result is
# overwritten by the load() below, so it is commented out. Uncomment it (and
# the save()) if you want to generate a fresh set of 1000 simulations.

#ant_sim = make.simmap( tree = ladderize(t_adults), x= adults.sp.sim , model = "ER", pi = 'fitzjohn',
#                     nsim = 1000)

#save(ant_sim, file = './evo.models/ant_sim.RData')

load(file = './evo.models/ant_sim.RData')


sum_sim = describe.simmap(ant_sim)
sum_sim

plot(density(ant_sim))

```

### NYMPHS AND ADULTS GENERA ANALYSES ####

The same process made to antennomere analysis made for adults species will be made for nymphs and adults at the genus level

## NYMPHS ####
```{r, fig.width = 10, fig.height = 15}
# filtering data, removing 0 and species out of Coreidae

nymph_data = gen_nymphs%>% 
  filter(Antennae != 0, 
         !Species_Name %in% c("Jadera_haematoloma", "Harmostes_serratus",
                              "Oedancala_sp.", "Oncopeltus_fasciatus",
                              "Largus_sp.", "Dysdercus_mimus",
                              "Dysdercus_suturellus", "Halyomorpha_halys")) %>% 
  mutate(Species_Name = case_when(
    Species_Name == "Holhymenia_tibialis" ~ "Holhymenia_tibialis / clavigera",
    TRUE ~ Species_Name
  ))

# Now we prune the tree to the data we have on Coreidae nymphs

t_nymphs = keep.tip(t1, nymph_data$Species_Name)


# Creating nymphs matrix so phytools can read our data

nymphs.gen.sim = setNames(as.factor(nymph_data$Sum), nymph_data$Species_Name)

nymphs.gen.sim

# Model Mk, all models

#fit.ER.equal.nymph = fitMk(ladderize(t_nymphs), nymphs.gen.sim, model = "ER",
#                   pi = 'equal')


#fit.ER.fitz.nymph= fitMk(ladderize(t_nymphs), nymphs.gen.sim , model = "ER",
#                    pi = 'fitzjohn')

#save(fit.ER.equal.nymph, file = './evo.models/er-equal-nymph.RData')
#save(fit.ER.fitz.nymph, file = './evo.models/er-fitz-nymph.RData')

load( file = './evo.models/er-equal-nymph.RData')
load( file = './evo.models/er-fitz-nymph.RData')

#fit.ARD.equal.nymph = fitMk(ladderize(t_nymphs), nymphs.gen.sim , model = "ARD",
#                            pi = 'equal')

#fit.ARD.fitz.nymph = fitMk(ladderize(t_nymphs), nymphs.gen.sim, model = "ARD",
#                           pi = 'fitzjohn')

#save(fit.ARD.equal.nymph, file = './evo.models/ARD-equal-nymph.RData')
#save(fit.ARD.fitz.nymph, file = './evo.models//ARD-fitz-nymph.RData')

load( file = './evo.models/ARD-equal-nymph.RData')
load( file = './evo.models/ARD-fitz-nymph.RData')

# Hidden Rates Model, HRM

#fitHRM.er.equal.nymph = fitHRM(ladderize(t_nymphs), nymphs.gen.sim , model = "ER",
#                       pi = 'equal')

#fitHRM.er.fitz.nymph = fitHRM(ladderize(t_nymphs), nymphs.gen.sim , model = "ER",
#                              pi = 'fitzjohn')

#save(fitHRM.er.equal.nymph, file = './evo.models/hrm-er-equal-nymph.RData')
#save(fitHRM.er.fitz.nymph, file = './evo.models/hrm-er-fitz-nymph.RData')

load(file = './evo.models/hrm-er-equal-nymph.RData')
load(file = './evo.models/hrm-er-fitz-nymph.RData')

#fitHRM.ard.equal.nymph = fitHRM(ladderize(t_nymphs), nymphs.gen.sim , model = "ARD",
#                                pi = 'equal')


#fitHRM.ard.fitz.nymph = fitHRM(ladderize(t_nymphs), nymphs.gen.sim , model = "ARD",
#                         pi = 'fitzjohn')

#save(fitHRM.ard.equal.nymph, file = './evo.models/hrm-equal-nymph.RData')
#save(fitHRM.ard.fitz.nymph, file = './evo.models/hrm-fitz-nymph.RData')


load(file = './evo.models/hrm-equal-nymph.RData')
load(file = './evo.models/hrm-fitz-nymph.RData')

# comparing models
fit_models_nymphs = anova(fit.ER.equal.nymph, fit.ARD.equal.nymph,
                       fitHRM.er.equal.nymph, fitHRM.ard.equal.nymph,
                       fit.ER.fitz.nymph, fit.ARD.fitz.nymph,
                       fitHRM.er.fitz.nymph, fitHRM.ard.fitz.nymph)

# The chosen method was fit.HRM.er.fitz.nymph once it had the lowe AIC and
# less degrees of freedom (d.f. = 1)

fitHRM.er.fitz.nymph
plot(fitHRM.er.fitz.nymph)

anc.nymph = ancr(fitHRM.er.fitz.nymph)

cols = setNames(c("#0E3768","blue", "#94129F", "#ea22fb",  
     "#DE627A", "pink", "#ce4326", "#ff8b5f"),
                fitHRM.er.fitz.nymph$states)


#pdf(file = './ancestral-nymph.pdf', h = 40, w = 20)
plot(anc.nymph,  
     args.plotTree = list(lwd = 3, fsize = 1, 
                          ftype = 'i',
                          outline = T,
                          cex = 0.2, offset = 3
     ),
     args.nodelabels = list(cex = 0.3, piecol = cols),
     args.tiplabels = list(cex = 0.3)
     
)
#dev.off()
```

```{r, fig.width = 15, fig.height = 10}
#ard_sim_nymph = make.simmap( tree = ladderize(t_nymphs), x= nymphs.gen.sim , model = "ER", pi = 'fitzjohn',
#                             nsim = 1000)

#save(ard_sim_nymph, file = './evo.models/ard_sim_nymph.RData')

load( file = './evo.models/ard_sim_nymph.RData')

sum_sim_nymph = describe.simmap(ard_sim_nymph)
sum_sim_nymph

plot(density(ard_sim_nymph))
```

## ADULTS - GENUS LEVEL ####

```{r}
# Now, we will use the adults data colapsed at genus level

adults_data2 = gen_adults %>% 
  filter(Antennae != 0, 
         !Species_Name %in% c("Jadera_haematoloma", "Harmostes_serratus",
                              "Oedancala_sp.", "Oncopeltus_fasciatus",
                              "Largus_sp.", "Dysdercus_mimus",
                              "Dysdercus_suturellus", "Halyomorpha_halys")) %>% 
  mutate(Species_Name = case_when(
    Species_Name == "Holhymenia_tibialis" ~ "Holhymenia_tibialis / clavigera",
    TRUE ~ Species_Name
  ))

# We need to make sure that only species present in nymphal data are present
# in the adults data, so we can compare them equally

adults_data2 = gen_adults %>%
  mutate(Species_Name = case_when(
    Species_Name == "Holhymenia_tibialis" ~ "Holhymenia_tibialis / clavigera",
    TRUE ~ Species_Name
  )) %>% 
  filter(Antennae != 0, 
         !Species_Name %in% c("Jadera_haematoloma", "Harmostes_serratus",
                              "Oedancala_sp.", "Oncopeltus_fasciatus",
                              "Largus_sp.", "Dysdercus_mimus",
                              "Dysdercus_suturellus", "Halyomorpha_halys"),
         Species_Name %in% nymph_data$Species_Name)
```

```{r, fig.width = 15, fig.height = 10}
# Now we prune the tree to the data we have on Coreidae adults genera

t_adults_gen = keep.tip(t1, adults_data2$Species_Name)

# adults matrix

adults.gen.sim = setNames(as.factor(adults_data2$Sum), adults_data2$Species_Name)
adults.gen.sim

# Mk model
#fit.ER.equal.ag = fitMk(ladderize(t_adults_gen), adults.gen.sim, model = "ER",
#           pi = 'equal')


#fit.ER.fitz.ag = fitMk(ladderize(t_adults_gen), adults.gen.sim , model = "ER",
#                 pi = 'fitzjohn')

#save(fit.ER.equal.ag, file = './evo.models/er-equal-adults-genus.RData')
#save(fit.ER.fitz.ag, file = './evo.models/er-fitz-adults-genus.RData')

load( file = './evo.models/er-equal-adults-genus.RData')
load( file = './evo.models/er-fitz-adults-genus.RData')

#fit.ARD.equal.ag = fitMk(ladderize(t_adults_gen), adults.gen.sim, model = "ARD",
#                        pi = 'equal')

#fit.ARD.fitz.ag = fitMk(ladderize(t_adults_gen), adults.gen.sim, model = "ARD",
#                       pi = 'fitzjohn')

#save(fit.ARD.equal.ag, file = './evo.models/ARD-equal-adults-genus.RData')
#save(fit.ARD.fitz.ag, file = './evo.models/ARD-fitz-adults-genus.RData')

load( file = './evo.models/ARD-equal-adults-genus.RData')
load( file = './evo.models/ARD-fitz-adults-genus.RData')

# Hidden rates model, HRM

#fitHRM.er.equal.ag = fitHRM(ladderize(t_adults_gen), adults.gen.sim, model = "ER",
#                           pi = 'equal')

#fitHRM.er.fitz.ag = fitHRM(ladderize(t_adults_gen), adults.gen.sim, model = "ER",
#                          pi = 'fitzjohn')

#save(fitHRM.er.equal.ag, file = './evo.models/hrm-er-equal-adults-genus.RData')
#save(fitHRM.er.fitz.ag, file = './evo.models/hrm-er-fitz-adults-genus.RData')

load(file = './evo.models/hrm-er-equal-adults-genus.RData')
load(file = './evo.models/hrm-er-fitz-adults-genus.RData')

#fitHRM.ard.equal.ag = fitHRM(ladderize(t_adults_gen), adults.gen.sim, model = "ARD",
#                            pi = 'equal')


#fitHRM.ard.fitz.ag = fitHRM(ladderize(t_adults_gen), adults.gen.sim, model = "ARD",
#                           pi = 'fitzjohn')

#save(fitHRM.ard.equal.ag, file = './evo.models/hrm-equal-adults-genus.RData')
#save(fitHRM.ard.fitz.ag, file = './evo.models/hrm-fitz-adults-genus.RData')


load(file = './evo.models/hrm-equal-adults-genus.RData')
load(file = './evo.models/hrm-fitz-adults-genus.RData')


# Comparing models
fit_models_ag = anova(fit.ER.equal.ag, fit.ARD.equal.ag,
                      fitHRM.er.equal.ag, fitHRM.ard.equal.ag,
                      fit.ER.fitz.ag, fit.ARD.fitz.ag,
                      fitHRM.er.fitz.ag, fitHRM.ard.fitz.ag)

# It was chosen the ER fitz model once it had the lowest AIC value and 
# 1 degree of freedom

fit.ER.fitz.ag
plot(fit.ER.fitz.ag)

```

## PLOTTING NYMPHS AND ADULTS GENERA TREE TOGETHER ####

```{r, fig.width = 15, fig.height = 10}
#Nymph 
anc.nymph = ancr(fitHRM.er.fitz.nymph, hide.hidden = F)

#Adults
anc.ag = ancr(fit.ER.fitz.ag)

tree_plot = fit.ER.fitz.ag$tree

anc.nymph$tree = tree_plot
anc.ag$tree = tree_plot

states = colnames(anc.nymph$ace)
cols = setNames(
  c("#0E3768","blue", "#94129F", "#ea22fb",  
    "#DE627A", "pink", "#ce4326", "#ff8b5f"),
  states
)

states2 = colnames(anc.ag$ace)
cols2 = setNames(
  c("#0E3768", "#94129F",
    "#DE627A",  "#ce4326"),
  states2
)
cols2

tip_state = as.character(nymphs.gen.sim[anc.nymph$tree$tip.label])
states2

tip.pies = sapply(
  states2,
  function(z) as.numeric(tip_state == z)
)

tip.pies = as.matrix(tip.pies)
rownames(tip.pies) = anc.nymph$tree$tip.label
colnames(tip.pies) = states2

#png(file = './figures/ancestral-facing.png', height = 400, width = 350,
#    units = "mm", res = 300)
#pdf(file = './figures/antennnomere_facing.pdf',
#    h = 12, w = 12)

layout(matrix(1:3, nrow = 1), widths = c(0.44, 0.12, 0.44))

plot(
  anc.nymph,
  args.plotTree = list(
    ftype = "off",
    lwd = 1.5,
    ylim = c(0, Ntip(tree_plot) + 1),
    mar = c(1, 0.5, 3, 0.5)
  ),
  args.nodelabels = list(
    piecol = cols,
    cex = 0.3
  )
)
par(fg = 'transparent')
tiplabels(
  pie = tip.pies,
  piecol = cols2,
  cex = 0.45,
  border = "transparent"
)
par(fg = 'black')
mtext("nymph", side = 3, line = 1, adj = 0)

## Middle: taxon labels, plotted once
pp <- get("last_plot.phylo", envir = .PlotPhyloEnv)

## Tips are nodes 1:Ntip(tree); arrange bottom to top
tip_order <- order(pp$yy[seq_len(Ntip(tree_plot))])

tip_labels <- tree_plot$tip.label[tip_order]
tip_y      <- pp$yy[tip_order]
y_lim      <- pp$y.lim

## Middle panel: use the plotted y coordinates, NOT 1:Ntip()
plot.new()
plot.window(xlim = c(-1, 1), ylim = y_lim)

text(
  x = 0,
  y = tip_y,
  labels = gsub("_", " ", tip_labels),
  cex = 1,
  font = 3
)


plot(
  anc.ag,
  args.plotTree = list(
    ftype = "off",
    direction = "leftwards",
    lwd = 1.5,
    ylim = c(0, Ntip(tree_plot) + 1),
    mar = c(1, 0.5, 3, 0.5)
  ),
  args.nodelabels = list(
    piecol = cols2,
    cex = 0.3
  ),
  args.tiplabels = list(
    cex = 0.45
  )
)
mtext("adults", side = 3, line = 1, adj = 1)

```


## SESSION INFO

```{r}
sessionInfo() %>% pander
```


